---
title: "MAT5317 Categorical Assignment 1"
author:
- Teng Li(7373086)
- Zhize Lu(300075114)
- Chutong Zhang(300311325)
output: html_document
header-includes:
- \renewcommand{\and}{\\}
- \usepackage{float}
- \floatplacement{figure}{H}
bibliography: References.bib
link-citations: yes
---


<style type="text/css">
body{
font-size: 12pt;
}
table{
font-size: 12pt;
}
</style>

```{r setup, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
library(plotly)
library(tidyverse)
library(kableExtra)
library(aplore3)
library(table1)
```

## Introduction

We were given the data set of The National Health and Nutrition Examination Survey (NHANES). The survey program has been conducted as a series of surveys designed to assess the health and nutritional status of adults and children in the United States since the 1960s, according to CDC [@CDC]. It combines in-person face-to-face interviews and physical examinations of participants for data collection.

The survey data wasn't a simple random sample, however. According to CDC's National Health and Nutrition Examination Survey: Plan and Operations, 1999--2010 [@SurveyDesign], the sampling strategy consists of several stages: 1. Selection of counties as primary sampling units (PSU). 2. selection of segments within PSUs that constitute blocks of households. 3. Selection of specific households within segments. 4. Selection of individuals within a household.

We aim to study the relationship between the weight variable and the other health related variables of the data.

## Method

We began our study by doing an exploratory analysis among the variables through various tables and charts. We then performed several hypothesis tests on some of the variables. Lastly we did a linear regression model fit to the response variable "weight" with other variables and confounders.

## Part 1: Exploratory Analysis

We began our analysis by giving a data dictionary of the data shown in Table 1 below. As one can see that some variables have a high percentage of missing values. In Part 2 we made hypothesis tests to decide if some of these variables could be excluded from the regression analysis in Part 3.

The weight variable was a continuous random variable in our data. A simple way of categorizing it was to consider the BMI indicator. As one could see there was an obese variable in the data. The weight variable was categorized by giving a threshold of 35 to the BMI value. A person is considered healthy if the BMI is below 35, and obese otherwise. Therefore, we used the obese variable as the categorical random variable in our project.

The  NHANES III data set from 2009-2010 is an ongoing and continuous survey focused on civilian noninstitutional population fo the United states and published by CDC every year, it is designed to be a surveillance of specific diseases and behaviors,providing statistical insights into the U.S population.[@SurveyDesign] In this assignment, we accessed the data set from the aplore3 package in R, which contains 6482 observations in 21 variables.Of these, 8(gender, marstat, vigwrk, modwrk, wlkbik, vigrecexr, and obesity)can be considered as categorical variables, while the rest as numerical variables.

In categorical variables, we need to specified the vigorous work or recreational activities and moderate work or recreational activities. vigorous work or recreational activities means that intense physical exertion leading to significant elevation in respiration and heart rate, it usually sustained for more than 10 minutes during the work or recreational activities. Moderate work and recreational activities means that activities require moderate physical effort and results in slight increase in respiration and heart rate, it usually sustained for less than 10 minutes during the work or recreational activities.[@CODE] 

In numerical variables, we also want to elucidate several variables related to cholesterol and blood pressure. Cholesterol is a type of lipid that helps body perform important body functions. Tchol indicates the Total Cholesterol, it is the total number of cholesterol circulation in people's blood. Hdl indicates the HDL-Cholesterol, it is the high-density lipoprotein cholesterol that helps people transfer excess cholestrol from the blood to the liver.[@CHOL] Sysbp stands for Systolic Blood Pressure and dbp stands for Diastolic Blood Pressure, systolic pressure repersents the maximum blood pressure when ventricles contract, while diastolic pressure denotes the minimum blood pressure registered just before the subsequent contraction.[@BP] We will discuss the thresholds for cholesterol and blood pressure in the following discussion. Two other variables in this dataset are worth noticing, psu and strata, it is the Pseudo-PSU and Pseudo-stratum used in sampling strategy. In this data set, stratum was defined by geography and psu were selected from every stratum with probability proportional to a measure of size(PPS). More details about survey weights for NHANES will be presented in the extra topic part.

In original data set, it exists 6482 observations and 37 of them are missing information for variable bmi. We deleted these missing data,only perform 6445 observations as total sample size and use BMI level to stratified observations, since the missing data less than 0.6% in total. As can be seen in table 1, the first three columns are different level of BMI based on CDC's classification,the last column is the overall performance for every variables. For categorical variables, first three column presented the total number of observation and percentage under the stratification condition, last column is the sum of the first three column and the percentage of the total observations. For numerical variables, first three columns are the mean and standard deviation under the stratification condition and the last column is the mean and standard deviation under overall level.
```{r}
#Table 1 
Bnhanes<-nhanes
Bnhanes<-mutate(Bnhanes,BMI_level = case_when(
    Bnhanes$bmi < 18.5 ~"Underweight",
    Bnhanes$bmi < 25 ~"Healthy weight",
    Bnhanes$bmi < 30 ~"Overweight",
    Bnhanes$bmi >= 30 ~"Obesity "))%>%
    filter(!is.na(Bnhanes$bmi))
  
label(Bnhanes$gender) <- "Gender"
label(Bnhanes$age)<-"Age"
label(Bnhanes$marstat)<-"Marital Status"
label(Bnhanes$samplewt)<-"Statistical Weight"
label(Bnhanes$psu)<-"Pseudo-PSU"
label(Bnhanes$strata)<-"Pseudo-stratum"
label(Bnhanes$tchol)<-"Total Cholesterol"
label(Bnhanes$hdl)<-"HDL-Cholesterol"
label(Bnhanes$sysbp)<-"Systolic Blood Pressure"
label(Bnhanes$dbp)<-"Diastolic Blood Pressure "
label(Bnhanes$wt)<-"Weight"
label(Bnhanes$ht)<-"Standing Height"
label(Bnhanes$bmi)<-"Body mass Index"
label(Bnhanes$vigwrk)<-"Vigorous Work Activity"
label(Bnhanes$modwrk)<-"Moderate Work Activity "
label(Bnhanes$wlkbik)<-"Walk or Bicycle "
label(Bnhanes$vigrecexr)<-"Vigorous Recreational Activities "
label(Bnhanes$modrecexr)<-"Moderate Recreational Activities"
label(Bnhanes$sedmin)<-"Minutes of Sedentary Activity per Week "
label(Bnhanes$obese)<-"Obese "

units(Bnhanes$age) <- "years"
units(Bnhanes$tchol) <- "mg/dL"
units(Bnhanes$hdl) <- "mg/dL"
units(Bnhanes$sysbp) <- "mm Hg"
units(Bnhanes$dbp) <- "mm Hg"
units(Bnhanes$wt) <- "Kg"
units(Bnhanes$ht) <- "cm"
units(Bnhanes$bmi) <- "Kg/m^2"
units(Bnhanes$sedmin) <- "mins"

my.render.cont <- function(x) {
    with(stats.apply.rounding(stats.default(x), digits=3), c("",
        "Mean (SD)"=sprintf("%s (± %s)", MEAN, SD)))
}

tbl<-table1 (~ gender+age+marstat+samplewt+psu+strata+tchol+hdl+sysbp+dbp+wt+ht+vigwrk+modwrk+wlkbik+vigrecexr+modrecexr+sedmin+obese |BMI_level,render.continuous=my.render.cont, data=Bnhanes)
kable(tbl,"html",caption="Table 1: Characteristics of the data set NHANES")%>%
 kable_styling(bootstrap_options = "striped", full_width = F)
```

According to CDC's classification on bodyweight, we have: BMI\<18.5 as Underweight, BMI between 18.5 and 24.9 as Health, BMI between 25 and 29.9 as Overweight, and BMI\>30 as obesity. We adopted this category and found that there was a slight positive relationship between bodyweight and the total cholesterol level. However, we noticed that there was a negative relationship between the HDL and bodyweight. Because of the fact that Tchol is the sum of HDL and LDL, we can conclude that the obese population has a high level of LDL and a low level HDL.

```{r}
nhanes%>%mutate(CDC_obese=case_when(
  bmi <18.5 ~ "Underweight", 
  bmi>=18.5 & bmi<=24.9 ~ "Healthy",
  bmi>24.9 & bmi<=29.9 ~ "Overweight",
  bmi>29.9  ~ "Obesity"
))%>%transmute(CDC_obese=factor(CDC_obese, levels=c("Underweight", "Healthy", "Overweight", "Obesity")), tchol, hdl)%>%na.omit()%>% group_by(CDC_obese)%>%summarise(Mean_tchol=mean(tchol), Mean_hdl=mean(hdl))%>%gather("Type", "Value",Mean_tchol:Mean_hdl)%>%
  plot_ly(x=~CDC_obese, y=~Value, color = ~Type, type = "scatter", mode="lines", height = 300,width = 400)%>%
  layout(
    title=list(text="Mean Cholesterol level across Categories of Bodyweight",font=list(size=12)),
    yaxis = list(range = c(45, 200),dtick = 50)
  )
```

According to ATPIII [@ATP], we can also categorize the cholesterol level.

This data set mainly focus on the observers between 16 to 80 years old. Among them, the average weight for male is greater than female among all ages, and as we can see from the line chart that the change in average weight with age follow the same trend across the gender, with a general tendency to sustained increase, followed by fluctuation and continuous decrease finally. It can be considered as there might exist some relationship with weight and age.
```{r}
#exploratory average weight in every age grouped by gender
gender<-nhanes%>%
  group_by(gender,age)%>%
  summarise(Avgwt=mean(wt,na.rm=TRUE), .groups="drop")
   ggplot(gender,aes(x=age, y=Avgwt), group=gender)+
     geom_point(aes(colour=gender),size=1 )+
     geom_line(aes(fill=gender,colour=gender,linetype=gender) )+
     geom_hline(yintercept = 86.2, linetype="dotted", size = 0.5)+
     geom_hline(yintercept = 74.9, linetype="dotted", size = 0.5)+
     annotate("text", y=86.2, x=18, label=" Male Avg.")+
     annotate("text", y=74.9, x=18, label="Female Avg.")+
      scale_y_continuous( labels = scales::comma  ) +
    theme(axis.text.x = element_text(size =7, vjust = 0.5, hjust = 0.5, angle = 45))+
   
   labs(title="Average Weight in Different Age",x="Age", y="Average weight (Kg)")+
   theme(plot.title = element_text(hjust = 0.5))
```

For the observations in different marital status, we also interested in the relationship between weight and marital status.The following box plot shows that the median weight under different marital status are all around 80 Kg, widowed observation have lowest weight among six categories. Married and Never Married observations have more people heavier than 130 Kg than other categories, which may not good for health.
```{r}
#box plot for weight VS marital status
Nhanes<-nhanes%>%
  na.omit()
ggplot(Nhanes, aes(x=marstat, y=wt, fill=marstat))+
        geom_boxplot() + 
        scale_fill_brewer(palette="Blues") +
        labs(title = "Plot for weight in each marital status", x="Marital Status", y="Weight (Kg)")+
        theme(axis.text.x = element_text(size =7, vjust = 0.5, hjust = 0.5, angle = 45),
              plot.title = element_text(hjust = 0.5)
       )
```

Different types of work and recreational activities also are interesting variables to discuss.We can see that only 1 missing data in these four variables,so we omit this missing data directly. For the vigorous activities, observations not in both work and recreational activities got the lowest and highest BMI. No matter the condition on work activities, upper quartile for observations are below 30 and lower quartile greater than 20 in observations have vigorous recreational activities. This means that majority people in this condition have a healthy BMI index. For moderate activities, observation not in both moderate work and recreational activities have the highest BMI and observation only in moderate work activities have the lowest BMI. No matter the condition on work activities, upper quartile for observations are slightly above 30 and lower quartile greater than 20 in observations have vigorous recreational activities. In general, we can see from two plots that observations have vigorous or moderate recreational activities trend to have healthy BMI index than others,and under same recreational activities condition observations in moderate or vigorous work activities have less number of observations have high BMI index. Hence, different types of work or recreational activities may have relationship with weight and affect BMI index in this way.
```{r} 
#boxplot for Vigorous activities VS bmi
Nhanes<-nhanes%>%
   mutate(vigrecexr = case_when(
    (nhanes$vigrecexr %in% "Yes") ~ "Vigorous Recreational Activities",
    (nhanes$vigrecexr %in% "No") ~ "Not Vigorous Recreational Activities"))%>%
   na.omit()
   ggplot(Nhanes, aes(x=vigwrk, y=bmi, fill=vigwrk))+
        geom_boxplot() + 
        facet_wrap(~vigrecexr) +
      geom_hline(yintercept = 18.5, linetype="dotted", size = 0.5)+
      geom_hline(yintercept = 30, linetype="dotted", size = 0.5)+
      annotate("text", y=18.5, x="Yes", label=" underweight")+
      annotate("text", y=30,x="Yes", label=" obsity")+
        scale_fill_brewer(palette="Blues") +
        labs(title = "Plot for BMI in Vigorous Activities", x=" Vigorous Work Activity", y="BMI")+
        theme(axis.text.x = element_text(size =7, vjust = 0.5, hjust = 0.5, angle = 45),
              plot.title = element_text(hjust = 0.5)
       )
  
  #boxplot for Moderate activities VS weight
Nhanes<-nhanes%>%
  mutate(modrecexr = case_when(
    (nhanes$modrecexr %in% "Yes") ~ "Moderate Recreational Activities",
    (nhanes$modrecexr %in% "No") ~ "Not Moderate Recreational Activities"))%>%
  na.omit()
  ggplot(Nhanes, aes(x=modwrk, y=bmi, fill=modwrk))+
        geom_boxplot() + 
       facet_wrap(~modrecexr) +
        scale_fill_brewer(palette="Blues") +
    geom_hline(yintercept = 18.5, linetype="dotted", size = 0.5)+
      geom_hline(yintercept = 30, linetype="dotted", size = 0.5)+
      annotate("text", y=18.5, x="Yes", label=" underweight")+
      annotate("text", y=30,x="Yes", label=" obsity")+
        labs(title = "Plot for BMI in Moderate Activities", x=" Moderate Work Activity", y="BMI")+
        theme(axis.text.x = element_text(size =7, vjust = 0.5, hjust = 0.5, angle = 45),
              plot.title = element_text(hjust = 0.5)
       )
```

## Part 2: Hypothesis Tests

```{r}
IndepTest<-setRefClass(
  "IndepTest",
  fields = list(data="data.frame", variable_x="character", variable_y="character",res="list"),
  methods=list(
    initialize=function(data, variable_x, variable_y){
      #make a contingency table for the variable X and Y
      for(x in variable_x){
        .self$res[[x]]<-data%>%select(all_of(c(x, variable_y)))%>%na.omit()%>%
          group_by(across(everything(), as.factor))%>%tally()%>%spread(key=variable_y, value=n)}
    },
    
    get_pvalue=function(){
      #do the chi-squared test
      pvalues<-map(res, function(x){
        y<-ungroup(x)%>%select(-any_of(group_vars(x)))%>%as.matrix()%>%chisq.test()
        return(y$p.value)
      })
      
      #format it nicely
      pvalues<-pvalues%>%data.frame(row.names = "p-value")%>%
        mutate(across(where(is.numeric), .fns = ~as.character(signif(.,4))))
      
      return(pvalues)}
  )
)
```

We first test the independence between obesity and marital status. We form the following contingency table:

```{r}
test_marstat<-IndepTest$new(nhanes, "marstat", "obese")
test_marstat$res$marstat%>%
  rename("\t" = marstat) %>%
  add_column(" " = c("Marital Status", rep(" ", nrow(.) - 1)), .before = "\t") %>%
  kable(caption = "Contingency Table")%>%
  kable_styling(bootstrap_options=c("striped","condensed"),position="float_left")%>%
  add_header_above(c(" " = 1, " " = 1, "Obesity" = 2))%>%
  column_spec(1:2,width = "10em",bold=TRUE)

pv<-test_marstat$get_pvalue()%>%as.numeric()
```

Let X be the categorical random variable for Marital Status and Y be the one for Obesity. Assuming a random sample of n trials. Define the count random variable $N_{ij}:=\sum_{k=1}^n \mathbf{I}_k(X=i, Y=j)$ where $\mathbf{I}_k$ is the indicator function for the k-th trial, then the joint random variables $[N_{11}, ..., N_{IJ}]$ has a Multinomial distribution $\vec{p}=[p_{11}, ..., p_{IJ}]$. Our hypothesis test is therefore:

```{=tex}
\begin{gather*}
H_0: p_{ij}= p_{i+} \cdot p_{+j} ~ \forall i,j\\
H_1:p_{ij} \neq p_{i+} \cdot p_{+j} ~ \forall i,j
\end{gather*}
```
We use the chi-squared test to conclude that there is not enough evidence to reject the null hypothesis with a p-value equal to `r pv`. In other words, we cannot conclude that there is a relationship between obesity and marital status.

We do the same test for other variables compared with obesity. From Table 2 we can see that we can reject the independence between obesity and wlkbik, vigrecexr and modrecexr variables.

```{r}
tests<-IndepTest$new(nhanes, c("vigwrk","modwrk","wlkbik","vigrecexr","modrecexr"), "obese")
tests$get_pvalue()%>%
  kable(caption = "p-values of Independence Tests between Different Variables and Obesity")%>%
  kable_styling(bootstrap_options=c("striped","condensed"),position="float_left")
```

Based on thresholds established by clinicians, we categorizing four variables related to the cholesterol and blood pressure level.

Cholesterol is an essential fat in the body. According to American Heart Association, We have: tchol of 240mg/dl or higher OR hdl under 40mg/dL as Dangerous Cholesterol level; tchol between 200-239mg/dL OR hdl between 40-59mg/dL for males OR between 50-59mg/dL for females as at risk level; tchol under 200mg/dL and hdl of 60mg/dL and higher as healthy level.

Blood pressure is typically categorized into different stages based on the systolic (top number) and diastolic (bottom number) readings. According to American Heart Association, We have: systolic level less than 120mm Hg and diastolic level less than 80mm Hg as normal blood pressure; systolic level between 120-129mm Hg and diastolic level less than 80mm Hg as elevated blood pressure; systolic level between 130-139mm Hg OR diastolic level between 80-89mm Hg as Hypertension Stage 1; systolic level 140mm Hg or higher OR diastolic level 90mm Hg or higher as Hypertension Stage 2; systolic level higher than 180mm Hg OR diastolic level higher than 120mm Hg as Hypertensive Crisis.

```{r}

nhanes_data <- nhanes

# Categorizing cholesterol levels
nhanes_data$chol_category <- with(nhanes_data, ifelse(
  tchol >= 240 | hdl < 40, "Dangerous",
  ifelse(
    tchol >= 200 & tchol < 240 | 
    (gender == "Male" & hdl >= 40 & hdl < 60) |  # Male and HDL between 40-59
    (gender == "Female" & hdl >= 50 & hdl < 60),   # Female and HDL between 50-59
    "At Risk",
    "Healthy"
  )
))

# Preview the updated variable
head(nhanes_data[, c("gender","tchol", "hdl", "chol_category")])

# Categorizing blood pressure
nhanes_data$bp_category <- with(nhanes_data, ifelse(
  sysbp < 120 & dbp < 80, "Normal",
  ifelse(sysbp >= 120 & sysbp <= 129 & dbp < 80, "Elevated",
  ifelse((sysbp >= 130 & sysbp <= 139) | (dbp >= 80 & dbp <= 89), "Hypertension Stage 1",
  ifelse(sysbp >= 140 | dbp >= 90, "Hypertension Stage 2",
  ifelse(sysbp > 180 | dbp > 120, "Hypertensive Crisis", NA))))))

# Preview the new variable
head(nhanes_data[, c("sysbp", "dbp", "bp_category")])
```

## Exrtra topic: Survey weights in complex survey designs

Most large-scale surveys often involve a combination of multiple sampling design techniques, like stratification and cluster sampling, enable researchers to obtain accurate estimates while catering to practical and cost considerations. A central tenet to these designs is the concept of sampling weights, pivotal in ensuring unbiased estimation.

Sampling weights are best defined as the inverse of the probability that a specific unit gets selected in the sample. These weights adjust for design-imposed inequalities in selection probabilities and are used tp compute point estimates.

In the case of the stratified random sampling. The population U of size N is partitioned into stratums denoted by $U_1,...,U_h,...,U_H$. The size of $h$th stratum is denoted by $N_h$.In stratum $h$, a random sample $S_h$ of size $n_h$ is selected based on a sampling design, here we use simple random sampling for simplicity and efficiency. In this stratified random sampling, the estimate of population total can be show as following[@Lohr_2022]:

$\hat{t}_{str}=\sum_{h=1}^H\sum_{j\in S_h}w_{hj}y_{hj}$

where the $w_{hj}=N_h/n_h$ represents the sampling weight of the $j$th observation in the $h$th stratum, $y_{hj}$. Note that the probability of sample selection of the $j$th unit in the $h$th stratum is $\pi_{hj}=n_h/N_h$.In this case, the sampling weight is the inverse of such probability $\pi_{hj}$. The unbiased estimator of the population mean $\bar{y}_U$ can also be shown with sampling weight as following[@Lohr_2022]:

$\hat{\bar{y}}_{str}=\frac{\sum_{h=1}^H\sum_{j\in S_h}w_{hj}y_{hj}}{\sum_{h=1}^H\sum_{j\in S_h}w_{hj}}$

Cluster sampling is another complex sampling technique for large-scale surveys when the population elements are dispersed and the the fieldwork is costly. We sample the primary sampling units(psu's) which are often from the natural groupings of the population elements. In cluster sampling, we have $N$ as the number of psu's in the population. $i$th psu contains $M_i$ elements. For sample of psu's, $S$, We denote $n$ as the number of psu's in the sample. For a two-stage cluster sampling, a $S_i$ sub-sample of secondary sampling units(ssu's) is chosen from $i$th psu, $i=1,...,n$. The sub-sample size is $m_i$.

In the case of two-stage cluster sampling with equal probabilities, the sampling weight can be expressed as the following: $w_{ij}=1/\pi_{ij}=\frac{NM_i}{nm_i}$, where $\pi_{ij}$ is the probability that the $j$th ssu in the $i$th psu is in the sample. For unequal probabilities, we need the probability that the $i$th psu is in the sample,$\pi_i$, and the probability that the $j$th ssu is in the sample given that the $i$th psu is in the sample, $\pi_{j|i}$. Then, the sampling weight is given by $w_{ij}=1/(\pi_i\pi_{j|i})$.

With the sampling weight shown above, the estimator of the population total in cluster sampling can also be show in the following forms[@Lohr_2022]:

$\hat{t}=\sum_{i\in S}\sum_{j\in S_i}w_{ij}y_{ij}$

and the estimator of the population mean:

$\hat{\bar{y}}=\frac{\sum_{i\in S}\sum_{j\in S_i}w_{ij}y_{ij}}{\sum_{i\in S}\sum_{j\in S_i}w_{ij}}$

In essence, sampling weights in complex survey designs play a crucial role in ensuring that the survey results are both accurate and generalizable to the broader population.














# Conclusion

# References

